---
title: "Dispersal Generator Test"
author: "July Pilowsky"
format: html
editor: visual
---

```{r setup, include = FALSE}
library(poems)
library(raster)
library(terra)
library(sf)
library(tidyverse)
library(paletteer)
library(cowplot)
library(landscapemetrics)
library(here)
library(qs)
library(tidyterra)
bbox <- extent(-128, -61, 15, 54)
proj_crs <- "+proj=aea +lon_0=-94.5 +lat_1=21.5 +lat_2=47.5 +lat_0=34.5 +datum=WGS84 +units=m +no_defs"
bioclim1940 <- rast("~/Documents/Very_Large_Data/ERA5/BioClim_1940.nc")
bioclim1940_proj <- project(bioclim1940, proj_crs)
template_raster <- rast("/Users/caryinstitute/Documents/Very_Large_Data/baseline/zip/1940AD_pop/uopp_1940AD.asc") %>% 
  project(bioclim1940_proj) %>% 
  crop(bioclim1940_proj)
```

# Overview

In the package `poems`, a dispersal generator can be used to calculate dispersal rates between all grid cells in a region based on a) dispersal capabilities of the organism, b) a *friction landscape*, and c) population density of the organism across cells. In this test of the dispersal generator for the house finch, I will simply test aspect a).

# Create Region Object

To do anything spatial in `poems`, you need a `Region` object. Regions define the study region: its extent, spatial resolution, and map projection. It helps to have a template raster (in this case, my raster of mean annual temperature for 1940.)

```{r template raster}
plot(template_raster)
```

My template raster looks great, except that it has islands, where I know there aren't any house finches. I'm going to remove these islands using the helpful package `landscapemetrics`.

```{r create region object}
template_raster[!is.na(template_raster)] <- 1
template_raster[is.na(template_raster)] <- 0
patch_raster <- template_raster %>% get_patches(class = 1) %>% 
  .[[1]] %>% .[[1]] %>% 
  raster::mask(raster(bioclim1940_proj[[13]]))
patch_raster[patch_raster != 23] <- NA
finch_region <- Region$new(template_raster = patch_raster)
plot(finch_region$region_raster)
```

# Create Dispersal Generator (and associated data)

The dispersal generator object needs three datasets to go with it.

1.  A lookup table to convert maximum dispersal distance to "dispersal b" (the value that actually goes into the dispersal function)
2.  A distance matrix that divides all possible dispersal distances into classes
3.  A table of distances from each grid cell to every other grid cell

It significantly speeds up operations if we have all of these three datasets ready to go along with the dispersal generator itself.

Here is the lookup table, based on a maximum dispersal distance of 1500 km according to the prior distributions I've calculated from the literature:

```{r b lookup}
b_lookup <- data.frame(d_max = -Inf, b = 0:904)
for (i in 2:904) {
  b_lookup$d_max[i] <- which.max(exp(-1*(1:1501)/b_lookup$b[i]) <= 0.19)
}
b_lookup$d_max[905] <- 1501
ggplot(b_lookup, aes(x = b, y = d_max)) + geom_line()
```

Here is the dispersal generator object:

```{r dispersal generator}
dispersal_gen <- DispersalGenerator$new(
  region = finch_region,
  dispersal_max_distance = 1500, # km
  distance_classes = seq(10, 1500, 10),
  distance_scale = 1000, # km
  dispersal_function_data = b_lookup,
  # dispersal_friction = DispersalFriction$new(conductance = friction),
  inputs = c("dispersal_p",
             "dispersal_r"),
  decimals = 3
)
```

I have commented out the friction landscape for now.

Now I will generate the distance matrix and the distance cell data and save them for later use.

```{r generate distance data}
# distance_matrix <- dispersal_gen$calculate_distance_matrix()
# dispersal_gen$calculate_distance_data(distance_matrix = distance_matrix)
# qs::qsave(distance_matrix, here("Data/Input/distance_matrix.qs"))
# qs::qsave(dispersal_gen$distance_data, here("Data/Input", "dispersal_distance_data.qs"))
distance_matrix <- qread(here("Data/Input", "distance_matrix.qs"))
distance_data <- qread(here("Data/Input", "dispersal_distance_data.qs"))
dispersal_gen$distance_data <- distance_data
```

# Generate Dispersal Rates

Here I will generate dispersals using some realistic values from my prior distributions.

```{r generate dispersals}
test_dispersal <- dispersal_gen$generate(input_values =
                                           list(dispersal_p = 0.5,
                                                dispersal_r = 500))$dispersal_data
dispersal_out <- test_dispersal[[1]] %>% group_by(source_pop) %>% 
  summarise(avg_dispersal_rate = mean(dispersal_rate)) %>% 
  right_join(rowid_to_column(finch_region$coordinates, "source_pop"))
dispersal_in <- test_dispersal[[1]] %>% group_by(target_pop) %>% 
  summarise(avg_dispersal_rate = mean(dispersal_rate)) %>% 
  right_join(rowid_to_column(finch_region$coordinates, "target_pop"))
ggplot(dispersal_out, aes(x, y, fill = avg_dispersal_rate)) +
  geom_tile() + 
  scale_fill_paletteer_c(`"pals::ocean.speed"`, 
                         name = "Average\nOutward\nDispersal",
                         limits = c(0, 0.012)) +
  theme_map()
ggplot(dispersal_in, aes(x, y, fill = avg_dispersal_rate)) +
  geom_tile() + 
  scale_fill_paletteer_c(`"pals::ocean.speed"`, 
                         name = "Average\nInward\nDispersal",
                         limits = c(0, 0.012)) +
  theme_map()
```

# Incorporate Friction Landscapes

I have created 3 friction landscapes: one with an effect of urbanization, one with an effect of elevation, and one with both multiplied. I want to see the effects these have on dispersal.

```{r urban friction dispersal}
urban_friction <- stack(here("Data/Input/urban_friction_landscape.grd"))
urban_friction_dispersal <- dispersal_gen$clone()
urban_friction_dispersal$dispersal_friction <- DispersalFriction$new(conductance = urban_friction)
# urban_friction_dispersal$calculate_distance_data(distance_matrix = distance_matrix)
urban_friction_dispersal$distance_data <- qread(here("Data/Input/urban_friction_distance.qs"))
test_dispersal <- urban_friction_dispersal$generate(input_values =
                                           list(dispersal_p = 0.5,
                                                dispersal_r = 500))$dispersal_data
dispersal_out <- test_dispersal[[1]] %>% group_by(source_pop) %>% 
  summarise(avg_dispersal_rate = mean(dispersal_rate)) %>% 
  right_join(rowid_to_column(finch_region$coordinates, "source_pop"))
dispersal_in <- test_dispersal[[1]] %>% group_by(target_pop) %>% 
  summarise(avg_dispersal_rate = mean(dispersal_rate)) %>% 
  right_join(rowid_to_column(finch_region$coordinates, "target_pop"))
ggplot(dispersal_out, aes(x, y, fill = avg_dispersal_rate)) +
  geom_tile() + 
  scale_fill_paletteer_c(`"pals::ocean.speed"`, 
                         name = "Average\nOutward\nDispersal",
                         limits = c(0, 0.012)) +
  theme_map()
ggplot(dispersal_in, aes(x, y, fill = avg_dispersal_rate)) +
  geom_tile() + 
  scale_fill_paletteer_c(`"pals::ocean.speed"`, 
                         name = "Average\nInward\nDispersal",
                         limits = c(0, 0.012)) +
  theme_map()
```

This friction landscape could clearly use some work because it has killed the dispersal dead.

Let's see if the elevation friction landscape does any better.

```{r elevation friction dispersal}
elevation_friction <- stack(here("Data/Input/elevation_friction_landscape.grd"))
elevation_friction_dispersal <- DispersalGenerator$new(
  region = finch_region,
  dispersal_max_distance = 1500, # km
  distance_classes = seq(10, 1500, 10),
  distance_scale = 1000, # km
  dispersal_function_data = b_lookup,
  dispersal_friction = DispersalFriction$new(conductance = elevation_friction),
  inputs = c("dispersal_p",
             "dispersal_r"),
  decimals = 3
)
# elevation_friction_dispersal$calculate_distance_data(distance_matrix = distance_matrix)
elevation_friction_dispersal$distance_data <- qread(here("Data/Input/elevation_friction_distance.qs"))
test_dispersal <- elevation_friction_dispersal$generate(input_values =
                                           list(dispersal_p = 0.5,
                                                dispersal_r = 500))$dispersal_data
dispersal_out <- test_dispersal[[1]] %>% group_by(source_pop) %>% 
  summarise(avg_dispersal_rate = mean(dispersal_rate)) %>% 
  right_join(rowid_to_column(finch_region$coordinates, "source_pop"))
dispersal_in <- test_dispersal[[1]] %>% group_by(target_pop) %>% 
  summarise(avg_dispersal_rate = mean(dispersal_rate)) %>% 
  right_join(rowid_to_column(finch_region$coordinates, "target_pop"))
ggplot(dispersal_out, aes(x, y, fill = avg_dispersal_rate)) +
  geom_tile() + 
  scale_fill_paletteer_c(`"pals::ocean.speed"`, 
                         name = "Average\nOutward\nDispersal",
                         limits = c(0, 0.012)) +
  theme_map()
ggplot(dispersal_in, aes(x, y, fill = avg_dispersal_rate)) +
  geom_tile() + 
  scale_fill_paletteer_c(`"pals::ocean.speed"`, 
                         name = "Average\nInward\nDispersal",
                         limits = c(0, 0.012)) +
  theme_map()
```

# Density-Dependent Dispersal

`poems` has four parameters for setting density-dependent dispersal, as follows:

1.  `dispersal_source_n_k`: Dispersal proportion (p) density dependence via source population abundance divided by carrying capacity (n/k), where p is reduced via a linear slope (defined by two list items) from n/k \<= *cutoff* (p = 0) to n/k \>= *threshold*.
2.  `dispersal_target_k`: Dispersal rate (r) density dependence via target population carrying capacity (k), where r is reduced via a linear slope (through the origin) when k \<= *threshold.*
3.  `dispersal_target_n`: Dispersal rate (r) density dependence via target population abundance (n), where r is reduced via a linear slope (defined by two list items) from n \>= *threshold* to n \<= *cutoff* (r = 0) or visa-versa.
4.  `dispersal_target_n_k`: Dispersal rate (r) density dependence via target population abundance divided by carrying capacity (n/k), where r is reduced via a linear slope (defined by two list items) from n/k \>= *threshold* to n/k \<= *cutoff* (r = 0) or visa-versa.

Now that I have data on habitat suitability, I can test out some of these. Let's look at probability of occurrence multiplied by a population density pulled from my literature review for that parameter.

```{r carrying capacity map}
carrying_capacity <- rast(here("Data/Input/habitat_suitability_v1.tif"))*200000
ggplot() +
  geom_spatraster(data = carrying_capacity, aes(fill = `1940`)) +
  scale_fill_paletteer_c("pals::linearl", name = "Carrying Capacity") +
  ggtitle("1940")
```

Let's try a test where the target habitat suitability for dispersal is the maximum carrying capacity (`dispersal_target_k` = 200000).

```{r test dispersal target k}
test_dispersal <- dispersal_gen$generate(input_values =
                                           list(dispersal_p = 0.5,
                                                dispersal_r = 500))$dispersal_data
# population_dispersal(
#       replicates = 10,
#       time_steps = 78,
#       years_per_step = 1,
#       populations = 7,
#       demographic_stochasticity = TRUE,
#       density_stages = c(0, 1, 1),
#       dispersal = dispersal_gen$dispersal_data,
#       dispersal_stages = c(0, 1, 0.5),
#       dispersal_source_n_k = list(cutoff = -0.5, threshold = 1.5),
#       dispersal_target_k = 5,
#       dispersal_target_n = list(threshold = 10, cutoff = 15),
#       simulator = simulator
#     )
```
