---
title: "Evaluating FeederWatch data on House Finch"
author: "July Pilowsky"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(vroom)
library(here)
library(mapdata)
library(cowplot)
library(paletteer)
map <- purrr::map
i_am("Scripts/Exploration/feederwatch.qmd")
source(here("Scripts/convenience_functions.R"))
```

# Introduction

My goal for this document is to visualize first sightings of House Finch (*Haemorhous mexicanus*) across North America according to FeederWatch. Of course to get proper first arrival times I would need to account for the Signor-Lipps effect, but this is simply a first pass. I will start with the 1996-2000 dataset.

```{r read in 1996 dataset, message = F}
pfw1996 <- vroom(here("Data/Validation/FeederWatch/PFW_1996_2000_public.csv")) %>% 
  ZeroFillPFW(SpeciesCodes = "houfin", rollup = F)
ggplot(pfw1996, aes(x = HOW_MANY)) + geom_histogram() + xlab("Number of Finches")

```

Clearly there are some unrealistic outliers, a problem with the dataset that I have seen pointed out in various papers analyzing this dataset. I am going to remove extreme outliers.

```{r better histogram}
pfw1996 %>% pull(HOW_MANY) %>% quantile(c(0.01, 0.99))
pfw1996 %>% filter(HOW_MANY < 42) %>% 
  mutate(finches_effort = HOW_MANY) %>% 
  ggplot(aes(x = finches_effort)) + geom_histogram() + 
  xlab("Number of House Finches")
```

Now I am going to group by latitude and longitude and sum how many finches were seen per lat-long in 1996, and plot this over the continent.

```{r 1996 finch map}
usa <- map_data("usa")
canada <- map_data("worldHires", "Canada")
pfw1996 %>% group_by(LATITUDE, LONGITUDE) %>% 
  filter(Year==1996) %>% 
  summarize(Finches = if_else(sum(HOW_MANY)>0, "Present", "Absent")) %>% 
  ggplot() +
  geom_polygon(data = usa, 
                                 aes(x=long, y = lat, group = group), 
                                 fill = "gray50", 
                                 color="black") +
    geom_polygon(data = canada, aes(x=long, y = lat, group = group), 
                 fill = "gray50", color="black") + 
    coord_quickmap(xlim = c(-140, -50), ylim = c(25, 55)) +
  geom_point(aes(x = LONGITUDE, y = LATITUDE, color = Finches), alpha = 0.6) +
  scale_color_paletteer_d("nbapalettes::huskies") + theme_map() +
  ggtitle("House Finch Observations in 1996")
  
```

I look at this map and I see that I can create a validation metric where I rasterize presence/absence of house finches to the spatial resolution of the model, and I require that the simulation be able to recreate the pattern of presence/absence of house finches in each year.

Now I'm going to create a dataset of house finches from 1988 to 1995, a period when the range of the House Finch was in its final stages of infilling, which is more useful for the validation metric.

```{r 1988 finch map, message = F}
pfw1988 <- vroom(here("Data/Validation/FeederWatch/PFW_1988_1995_public.csv")) %>% 
  ZeroFillPFW(SpeciesCodes = "houfin", rollup = F) %>% 
  filter(HOW_MANY < 70)
pfw1988 %>% group_by(LATITUDE, LONGITUDE) %>% 
  filter(Year==1988) %>% 
  summarize(Finches = if_else(sum(HOW_MANY)>0, "Present", "Absent")) %>% 
  ggplot() +
  geom_polygon(data = usa, 
               aes(x=long, y = lat, group = group), 
               fill = "gray50", 
               color="black") +
  geom_polygon(data = canada, aes(x=long, y = lat, group = group), 
               fill = "gray50", color="black") + 
  coord_quickmap(xlim = c(-140, -50), ylim = c(25, 55)) +
  geom_point(aes(x = LONGITUDE, y = LATITUDE, color = Finches), alpha = 0.6) +
  scale_color_paletteer_d("nbapalettes::huskies") + theme_map() +
  ggtitle("House Finch Observations in 1988")
```

The difference between the 1988 and 1996 maps is clear: the range has infilled in the Midwest / Great Plains. This is a useful validation metric.

# Comparison

I want to compare abundance trends in FeederWatch against abundance trends from Breeding Bird Survey to see how well they correspond with each other.

```{r abundance trend comparison}
routes <- read_csv(here("Data/Input/Routes.csv"), col_types = "iicciddiiii") %>% 
  mutate(ID = str_c(CountryNum, StateNum, Route))
weather <- vroom(here("Data/Input/Weather.zip"), col_types = "iiiciiiidiiiciiiicccii")
hofi <- here("Data/Input/States") %>% list.files(full.names = T) %>% 
  map(vroom, col_types = "iiiciiciiiii") %>% 
  bind_rows() %>% 
  filter(AOU=="05190") %>% 
  left_join(routes) %>% 
  left_join(weather) %>% 
  filter(QualityCurrentID==1)
trends1 <- hofi %>% filter(Year>1988, Year<1994) %>% 
  group_by(CountryNum, StateNum, Route) %>% 
  mutate(count = n(), sum = sum(SpeciesTotal)) %>% 
  filter(count==5, sum>0) %>% 
  nest() %>% 
  mutate(model = map(data, ~lm(SpeciesTotal ~ Year, data = .))) %>% 
  mutate(tidy = map(model, broom::tidy)) %>% 
  unnest(tidy) %>% 
  filter(term=="Year") %>% 
  select(-data) %>% 
  left_join(routes) %>% 
  filter(estimate>-9.142, estimate<19.942) %>% # remove extreme outliers
  mutate(Dataset = "BBS", Period = "1989-1993") %>% 
  ungroup()
trends2 <- hofi %>% filter(Year>1993, Year<1999) %>% 
  group_by(CountryNum, StateNum, Route) %>% 
  mutate(count = n(), sum = sum(SpeciesTotal)) %>% 
  filter(count==5, sum>0) %>% 
  nest() %>% 
  mutate(model = map(data, ~lm(SpeciesTotal ~ Year, data = .))) %>% 
  mutate(tidy = map(model, broom::tidy)) %>% 
  unnest(tidy) %>% 
  filter(term=="Year") %>% 
  select(-data) %>% 
  left_join(routes) %>% 
  filter(estimate>-12.66, estimate<9.45) %>% # remove extreme outliers
  mutate(Dataset = "BBS", Period = "1994-1998") %>% 
  ungroup()
trends3 <- pfw1988 %>% filter(Year>1988, Year<1994) %>% 
  group_by(LOC_ID) %>% mutate(years = n_distinct(Year), sum = sum(HOW_MANY)) %>% 
  filter(years==5, sum>0) %>% 
  nest() %>%
  mutate(model = map(data, ~lm(HOW_MANY ~ Year, data = .))) %>%
  mutate(tidy = map(model, broom::tidy)) %>%
  unnest(tidy) %>%
  filter(term=="Year") %>%
  mutate(Latitude = data[[1]] %>% pull(LATITUDE) %>% first(),
         Longitude = data[[1]] %>% pull(LONGITUDE) %>% first()) %>%
  filter(estimate>-5.25, estimate<7.8) %>% # remove extreme outliers
  mutate(Dataset = "FW", Period = "1989-1993") %>% 
  ungroup()
trends4 <- bind_rows(pfw1988, pfw1996) %>% 
  filter(Year>1993, Year<1999) %>% 
  group_by(LOC_ID) %>% mutate(years = n_distinct(Year), sum = sum(HOW_MANY)) %>% 
  filter(years==5, sum>0) %>% 
  nest() %>%
  mutate(model = map(data, ~lm(HOW_MANY ~ Year, data = .))) %>%
  mutate(tidy = map(model, broom::tidy)) %>%
  unnest(tidy) %>%
  filter(term=="Year") %>%
  mutate(Latitude = data[[1]] %>% pull(LATITUDE) %>% first(),
         Longitude = data[[1]] %>% pull(LONGITUDE) %>% first()) %>% 
  filter(estimate>-7.697, estimate<4.040) %>% #remove extreme outliers
  mutate(Dataset = "FW", Period = "1994-1998") %>% 
  ungroup()
list(trends1, trends2, trends3, trends4) %>% 
  map(~select(., Latitude, Longitude, Trend = estimate, Dataset, Period)) %>% 
  bind_rows() %>% 
  ggplot() +
  geom_polygon(data = usa, 
               aes(x=long, y = lat, group = group), 
               fill = "gray50", 
               color="black") +
  geom_polygon(data = canada, aes(x=long, y = lat, group = group), 
               fill = "gray50", color="black") + 
  coord_quickmap(xlim = c(-140, -50), ylim = c(25, 55)) +
  geom_point(aes(x = Longitude, y = Latitude, col = Trend),
             size = 2, alpha = 0.8) +
  theme_map() +
  scale_color_paletteer_c(`"ggthemes::Red-Blue-White Diverging"`, 
                          rescaler = scales::rescale_mid) +
  facet_grid(Dataset ~ Period)
```
