---
title: "mgsim: Round 2d Analysis"
author: "July Pilowsky"
format: html
editor: visual
toc: true
code-fold: true
lightbox: true
theme: materia
---

```{r setup, message = FALSE, warning = FALSE}
library(tidyplots)
library(paletteer)
library(here)
library(cowplot)
library(GGally)
library(abc)
library(tidyverse)
library(furrr)
library(bayestestR)
library(qs2)
library(epizootic)
plan(multicore)
source(here("Scripts/validation_metric_functions.R"))
source(here("Scripts/convenience_functions.R"))
source(here("Scripts/animate_sim.R"))
set_trust_promises(TRUE)
region <- qread(here("Data/Input/finch_region.qs"))
round2_metrics <- read_csv(here("Data/Validation/round2d_validation_metrics.csv"))
trend_metrics2 <- read_csv(here("Data/Validation/round2d_trend_metrics.csv")) |>
    pivot_longer(everything(), names_to = c("metric", "sim"), names_sep = "\\.\\.\\.")
round2_metrics$hm_trend_1970[round2_metrics$dc] <- trend_metrics2 |> filter(metric == "penalty1970") |> pull(value)
round2_metrics$hm_trend_1993[round2_metrics$dc] <- trend_metrics2 |> filter(metric == "penalty1993") |> pull(value)
round1_priors <- read_csv(here("Data/Input/sample_data_round1.csv"))
round1_metrics <- read_csv(here("Data/Validation/round1_validation_metrics.csv")) |>
    mutate(hm_trend_1970 = rep(NA_real_), hm_trend_1993 = rep(NA_real_))
round1_metrics$hm_trend_1970[round1_metrics$dc] <- round1_metrics$hm_trend[seq(1, 9752, 2)]
round1_metrics$hm_trend_1993[round1_metrics$dc] <- round1_metrics$hm_trend[seq(2, 9752, 2)]
round2_priors <- read_csv(here("Data/Input/sample_data_round2b.csv")) |>
  mutate(mortality_Sa_summer = round1_priors$mortality_Sa_summer,
         mortality_Ra_summer = mortality_Sa_summer)
```

# What is my goal in Round 2d of simulation?

I want to see if I can improve on previous rounds of simulation by using every available validation metric instead of narrowing to just two.

# Validation Metric Performance

## Presence in DC

To start with, let's look at presence of finches in DC. In the first round, 4876 simulations out of 10000 simulated the presence of house finches in Washington, DC in 1994, and in Round 2b, it was 4247 out of 10000.

```{r presence_dc}
sum(round2_metrics$dc)
```

Now the number of simulations with finches in DC has increased greatly to 8,988. I am a little nervous about this because it likely means that finch dispersal is much too aggressive.

## Other Validation Metrics

Moving on I can look at the other validation metrics:

```{r validation_metrics_plot, warning = FALSE}
validation_metrics <- bind_rows(round1_metrics, round2_metrics) |>
                      mutate(Round = factor(c(rep("1", 10000), rep("2", 10000)))) |>
                      filter(dc) |>
                      select(-hm_trend, -index, -dc)
quantiles <- validation_metrics[,1:7] |>
  map_dbl(\(x) quantile(x, prob = c(0.001), na.rm = T))
labels <- c("Mycoplasma presence", "Finch pres/abs",
            "Point prevalence", "Mycoplasma arrival", "Finch arrival",
            "Finch trend from 1970", "Finch trend from 1993")
colorpal <- paletteer_d("PNWColors::Sunset2")[c(1,3)]
plotlist <- pmap(list(names(validation_metrics)[1:7], quantiles, labels), function(x, y, z) {
    outliers_removed <- validation_metrics[, c(x, "Round")][pull(validation_metrics, x) < quantile(pull(validation_metrics, x), 0.95, na.rm=T), ]
    ggplot(data = outliers_removed, aes(x = .data[[x]], fill = Round, group = Round)) +
    geom_histogram(position = "dodge", bins = 30) +
    geom_vline(xintercept = y, color = "blue") +
    xlab(z) +
    scale_fill_manual(values = colorpal)
})
plot_grid(plotlist = plotlist)
```

This is a real mixed bag. Some metrics have improved a lot (finch trend from 1993). Some have gotten worse (Mycoplasma presence, Mycoplasma arrival). Others remained about the same (finch arrival, finch presence/absence, spatiotemporal point prevalence.)

# Selected models

Let's see if there's any improvement in selected models based on all the metrics. The previous round did not do very well because dispersal was much too aggressive, but perhaps this can be moderated.

```{r abc_everything, message = F, warning = F}
abc31 <- abc(param = round2_priors[round2_metrics$dc,],
             sumstat = round2_metrics |> filter(dc) |> select(3:9),
              target = rep(0, 7),
             tol = 0.01,
             transf = "log",
             method = "neuralnet")
qsave(abc31, here("Data/Validation/abc_round2d.qs"))
head(abc31$ss, 10)
selected_samples2 <- round2_priors |>
  rowid_to_column("sample") |>
  semi_join(abc31$unadj.values, copy = TRUE) |>
  pull(sample)
write_sftp_transfer("/glade/work/pilowskyj/Round2d",
                      "/Users/julypilowsky/Documents/mgsim/Data/Output/Round2d",
                      selected_samples2,
                      here("Scripts/Globus/abc_round2d.txt"))
ensemble_sa <- ensemble_mean(selected_samples2, abc31$weights, "Sa", here("Data/Output/Round2d"))
ensemble_ra <- ensemble_mean(selected_samples2, abc31$weights, "Ra", here("Data/Output/Round2d"))
ensemble_sj <- ensemble_mean(selected_samples2, abc31$weights, "Sj", here("Data/Output/Round2d"))
ensemble_rj <- ensemble_mean(selected_samples2, abc31$weights, "Rj", here("Data/Output/Round2d"))
ensemble_I1a <- ensemble_mean(selected_samples2, abc31$weights, "I1a", here("Data/Output/Round2d"))
ensemble_I2a <- ensemble_mean(selected_samples2, abc31$weights, "I2a", here("Data/Output/Round2d"))
ensemble_I1j <- ensemble_mean(selected_samples2, abc31$weights, "I1j", here("Data/Output/Round2d"))
ensemble_I2j <- ensemble_mean(selected_samples2, abc31$weights, "I2j", here("Data/Output/Round2d"))
total_population_size <- ensemble_sa + ensemble_ra + ensemble_sj + ensemble_rj + ensemble_I1a + ensemble_I2a + ensemble_I1j + ensemble_I2j
animate_sim(total_population_size, region, years = 1940:2016)
```

Once again, finch dispersal is much too rapid.